---
title: "Gold_Project"
author: "James DiLuca"
date: "2025-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Load in Gold Data
```{r}
library(readxl)
library(dplyr)
library(forecast)
library(ggplot2)
library(plotly)

file_path <- "C:/Users/james/Downloads/Gold_Prices.xlsx"

sheet_names <- c('Yearly_Avg', 'Quarterly_Avg', 'Monthly_Avg')

# Read each sheet into a separate data frame
Yearly_Avg <- read_excel(file_path, sheet = sheet_names[1]) %>%
  rename(Date = 1)

Quarterly_Avg <- read_excel(file_path, sheet = sheet_names[2]) %>%
  rename(Date = 1)

Monthly_Avg <- read_excel(file_path, sheet = sheet_names[3]) %>%
  rename(Date = 1)

```

# Clean Data
```{r}
# Convert to Just US Currency
Yearly_Avg_USD <- Yearly_Avg[, 1:2]
Quarterly_Avg_USD <- Quarterly_Avg[, 1:2] 
Monthly_Avg_USD <- Monthly_Avg[, 1:2]

```

# Create Timeseries for Monthly data
```{r}

Gold.ts <- ts(Monthly_Avg_USD$USD, start = c(1978, 1), end = c(2025, 1), freq = 12)

# Plot
plot(Gold.ts, main = "Monthly Average USD Time Series",
     xlab = "Year", ylab = "USD",
     col = "blue", lwd = 2)

```


# Set Test & Validation Data

```{r}

nValid <- 113


nTrain <- length(Gold.ts) - nValid

train.ts<- window(Gold.ts, end = c(2015,8))

valid.ts <- window(Gold.ts, start=c(2015, 9), end = c(2025, 1))
```

## ACF Plot

```{r}
diff.gold.ts <- diff(Gold.ts)
Acf(diff.gold.ts)

```

## Difference Plot

```{r}
plot.ts(diff(Monthly_Avg_USD$USD))

```



## Models 


### Linear Trend

```{r}

linear_mod <- tslm(train.ts ~ trend)
summary(linear_mod)

#print(linear_mod$fitted.values)

linear_mod_pred <- forecast(linear_mod, h= nValid, level = 90)
#print(linear_mod_pred$mean)

# ggplot2 + forecast package 
autoplot(linear_mod_pred) +
  autolayer(linear_mod_pred$mean, series ="Linear Forecast") +
  autolayer(linear_mod$fitted.values, series = "Linear Fit") + 
  autolayer(valid.ts, series = "Observed") + 
  xlab("Year") + ylab("Gold Price Per Ounce") + 
  guides(color = guide_legend(title = ""))


```


```{r}
linear.res <- linear_mod$residuals
acf(linear.res, lag.max = 12)
```


## Exponential Trend

```{r}

exp_mod <- tslm(train.ts ~ trend , lambda = 0) #lambda = 0 indicates exponential trend
summary(exp_mod)
exp_mod_pred <- forecast(exp_mod, h= nValid, level=0)

autoplot(exp_mod_pred) +
  autolayer(exp_mod_pred$mean, series="Exponential Forecast") + 
  autolayer(exp_mod$fitted.values, series = 'Exponential Fit') +
  autolayer(valid.ts, series = "Observed") + 
  xlab("Year") + ylab("Gold Price Per Ounce") + 
  guides(color = guide_legend(title = ""))


```

```{r}
exp.res <- exp_mod$residuals
acf(exp.res, lag.max = 12)
```

## Quadratic Trend Model

```{r}
quad_mod <- tslm(train.ts ~ trend + I(trend^2))
summary(quad_mod)
quad_mod_pred <- forecast(quad_mod, h= nValid, level=90)

Quad_mod_plot <- autoplot(quad_mod_pred) +
  autolayer(quad_mod_pred$mean, series ="Quadratic Forecast") +
  autolayer(quad_mod$fitted.values, series = "Quadratic Fit") +
  autolayer(valid.ts, series = "Observed") + 
  xlab("Year") + ylab("Gold Price Per Ounce") + 
  guides(color = guide_legend(title = ""))

plotly::ggplotly(Quad_mod_plot)

```


## Residuals

```{r}
quad.res <- quad_mod$residuals
acf(quad.res, lag.max = 12)
```

## AR1 on Residuals to Improve Quadratic Trend 

```{r}
train.res.ar.1 <- Arima(quad.res, order = c(1, 0, 0))


summary(train.res.ar.1)

quad.res.for <- forecast(quad.res, h =nValid, level = 90)

autoplot(quad.res)+
  autolayer(quad.res.for$fitted)
```



```{r}

autoplot(quad_mod_pred)+
  autolayer(quad.res.for$mean + quad_mod_pred$mean, series = "Improved Quad. Trend")+
  autolayer(valid.ts, series = "Observed") + 
  xlab("Year") + ylab("Gold Price Per Ounce") + 
  guides(color = guide_legend(title = ""))


```




```{r}
acf(train.res.ar.1$residuals, lag.max = 12)

```


## Additive Seasonality

```{r}
season_mod <- tslm(train.ts ~ season)
summary(season_mod)
season_mod_pred <- forecast(season_mod, h= nValid, level=0)


autoplot(season_mod_pred) +
  autolayer(season_mod_pred$mean, series ="Seasonality Forecast") +
  autolayer(valid.ts, series = "Observed") + 
  autolayer(season_mod$fitted.values, series = "Seasonality Fit") +
  guides(color = guide_legend(title = ""))
```



```{r}
season.res <- season_mod$residuals
acf(season.res, lag.max = 12)
```


```{r}
gold_reg_plot <- autoplot(season_mod_pred) +
  autolayer(season_mod_pred$mean, series ="Seasonality Forecast") +
  autolayer(quad_mod_pred$mean, series ="Quadratic Forecast") +
  autolayer(exp_mod_pred$mean, series="Exponential Forecast") +
  autolayer(linear_mod_pred$mean, series = "Linear Forecast") + 
  autolayer(quad_mod_pred$fitted, series = "Quadratic Fit")+
  autolayer(valid.ts, series = "Observed") + 
  labs(title = "Regression Based Model Forecasts", x = "Year", y = "Gold Price Per Ounce") + 
  guides(color = guide_legend(title = ""))

plotly::ggplotly(gold_reg_plot)

```


## Residual Plot


```{r}
autoplot(season_mod_pred$residuals) + 
  autolayer(season_mod_pred$residuals, series = "Training residuals") + 
  autolayer(valid.ts - season_mod_pred$mean, series = "Validation residuals")
```

##Holt Winters ZZZ Model

```{r}

gold.ets <- ets(train.ts, model = "ZZZ")

gold.ets.forecast <- forecast(gold.ets, h = nValid, level = 0)

autoplot(gold.ets.forecast) + 
  autolayer(gold.ets.forecast$mean, series = "ETS Forecast")+
  autolayer(valid.ts, series = "Observed")+
  guides(color = guide_legend(title = ""))
```


```{r}
ets.res <- gold.ets$residuals
acf(ets.res, lag.max = 12)
```


### Auto Arima

```{r}

gold.arima <- auto.arima(train.ts)
summary(gold.arima)
gold.arima.forecast <- forecast(gold.arima, h = nValid, level = 0)

gold_arima_plot <- autoplot(gold.arima.forecast)+
  autolayer(valid.ts, series = "Observed")+
  autolayer(gold.arima.forecast$mean, series ="ARIMA Forecast") +
  xlab("Time") + ylab("Gold Price per Ounce") +
  guides(color = guide_legend(title = ""))

plotly::ggplotly(gold_arima_plot)



```


```{r}
train.res.from.arima <- gold.arima$residuals

arima_res <- diff(train.res.from.arima)

acf(arima_res, lag.max = 12)


```

## NN model 

```{r}

p <- 10 # Number of previous time steps used for forecast
  
P <- 8 # Number of previous seasonal values to use
  
size <- 7 # Number of hidden nodes



gold.nnetar <- nnetar(train.ts, repeats =20, p = p, P = P, size = size)

gold.nnetar.forecast <- forecast(gold.nnetar, h = nValid)

autoplot(gold.nnetar.forecast) +
  autolayer(valid.ts, series = "Observed")+
  autolayer(gold.nnetar.forecast$mean, series ="NN Forecast") +
  autolayer(gold.nnetar.forecast$fitted, series = "NN Fit")+
  guides(color = guide_legend(title = ""))

```



## Naive


```{r}
gold.naive <- naive(train.ts, h = nValid, level = 0)

gold_naive_mod_pred <- forecast(gold.naive, h = nValid)

gold_naive <- autoplot(gold_naive_mod_pred) +
  autolayer(valid.ts, series = "Observed")+
  autolayer(gold_naive_mod_pred$mean, series ="Naive Forecast") +
  xlab("Time") + ylab("Gold Price per Ounce") +
  guides(color = guide_legend(title = ""))

plotly::ggplotly(gold_naive)
```


## Holt Winters 

```{r}

gold.holt.mod <- ets(train.ts, model = "AAN")
summary(gold.holt.mod)

gold.holt.pred <- forecast(gold.holt.mod, h = nValid, level = 0)

autoplot(gold.holt.pred)+
  autolayer(valid.ts, series = "Observed")+
  autolayer(gold.holt.pred$mean, series = "Holt Winter Forecast") +
  guides(color = guide_legend(title = ""))

```




```{r}
holt.res <- gold.holt.mod$residuals
acf(holt.res, lag.max = 12)
```






```{r}
dual_holt_plot <- autoplot(gold.holt.pred)+
  autolayer(valid.ts, series = "Observed")+
  autolayer(gold.holt.pred$mean, series = "Holt Winter (AAN) Forecast") +
  autolayer(gold.ets.forecast$mean, series = "Holt Winter (ZZZ) Forecast" )+
  labs(title = "Forecasts of Holt Winters Models", x = "Year", y = "Price of Gold per Ounce")+
  guides(color = guide_legend(title = ""))

plotly:: ggplotly(dual_holt_plot)

```

## Moving Average

```{r}
library(zoo)

ma.trailing <- rollmean(train.ts, k = 12, align = "right")
last.ma <- tail(ma.trailing, 1)

ma.trailing.pred <- ts(rep(last.ma, nValid), start = c(1978, nTrain + 1), end = c(1978, nTrain + nValid), freq = 12)

ma_plot <- autoplot(Gold.ts)+
  autolayer(ma.trailing, series = "Moving Average") +
  autolayer(ma.trailing.pred, series = "MA Forecast")+
  xlab("Time") + ylab("Gold Price per Ounce") +
  ggtitle("Moving Average Forecast")

plotly::ggplotly(ma_plot)

```


## Polynomial Performance 

```{r}
# Model 1: Linear Trend Model (linear_mod)
linear_mod_acc <- accuracy(linear_mod_pred$fitted, train.ts)
print(linear_mod_acc)
# Model 2: Exponential Trend Model (exp_model)
exp_mod_acc <- accuracy(exp_mod_pred$fitted, train.ts)
print(exp_mod_acc)
# Model 3: Quadratic Trend Model (quad_mod)
quad_mod_acc <- accuracy(quad_mod_pred$fitted, train.ts)
print(quad_mod_acc)
# Model 4: Additive Seasonality (season_mod)
season_mod_acc <- accuracy(season_mod_pred$fitted, train.ts)
print(season_mod_acc)
```

## Improved Quadratic Trend With ARIMA

```{r}

quad_mod_improved <- accuracy(quad.res.for$mean + quad_mod_pred$mean, valid.ts)
print(quad_mod_improved)

```



##Naive Model Performance

```{r}

naive_mod_acc <- accuracy(gold_naive_mod_pred$mean, valid.ts)
print(naive_mod_acc)
```

## ARIMA Model Performance:

```{r}
arima_mod_acc <- accuracy(gold.arima.forecast$mean, valid.ts)
print(arima_mod_acc)
```

## Holt Winter Model Performance:

```{r}

gold_holt_acc <- accuracy(gold.holt.pred$mean, valid.ts)
print(gold_holt_acc)

```

## Neural Net Model Performance 

```{r}

nn_acc <- accuracy(gold.nnetar.forecast$mean, valid.ts)
print(nn_acc)
```

## ETS ZZZ Model Performance

```{r}

ets_acc <- accuracy(gold.ets.forecast$mean, valid.ts)
print(ets_acc)
```

## Moving Average

```{r}
ma_acc <- accuracy(ma.trailing.pred, valid.ts)
print(ma_acc)
```

## Forecasting Using Improved Quadratic Model on Entire Dataset: 

```{r}

gold.quad <- tslm(Gold.ts ~ trend + I(trend^2))

gold_quad_pred <- forecast(gold.quad, h= nValid, level=95)

final.res <- gold.quad$residuals
  
quad.res.final <- forecast(final.res, h =nValid, level = 95)

gold_quad_plot <- autoplot(gold_quad_pred) +
  autolayer(quad.res.final$mean + gold_quad_pred$mean, series ="Quadratic Forecast") +
  labs(title = "Quadratic Forecast Improved Full", x = "Year", y = "Gold Price Per Ounce") + 
  guides(color = guide_legend(title = ""))

plotly::ggplotly(gold_quad_plot)

```

